\bfseries\LARGE\singlespacing{Spatiotemporal risks of \emph{Mycobacterium tuberculosis} transmission: A modelling study of environmental, clinical, and patient movement data}

### 2023-08-14 doc/paper/manuscript.tex

% the Wells-Riley model
The traditional Wells-Riley transmission model\cite{Riley1978AJE} is frequently used to estimate the risk of airborne transmission in a variety of indoor settings\cite{Andrews2014JID,Taylor2016IJTLD,Hella2017JInfect}, including primary care clinics\cite{Zurcher2022JID,McCreesh2021BMJGlobalHealth}. The Wells-Riley model assumes a well mixed airspace, where the risk of infection is modeled as a function of the number of infectious individuals in space, the generation rate of infectious doses (called quanta), the breathing rate per person, and the outdoor air supply rate. The quanta generation rate (commonly denoted with the parameter $q$) is unknown, but can be estimated by computing the risk of infection as the ratio of diseased/susceptible cases, and then solving the Wells-Riley equation for the parameter $q$\cite{Nardell1991ARRD,Escombe2008PLoSMed}. The Wells-Riley model considers the importance of ventilation by which infectious quanta are removed from the indoor space through outdoor air exchange. While the ventilation rate is difficult to estimate, an alternative formulation of the Wells-Riley equation exists that uses indoor CO$_2$ levels as a marker of exhaled-breath exposure\cite{Rudnick2003IndoorAir}, which is related to the ventilation rate.

@@

We used an anonymous personal movement tracking system (Xovis, Zollikofen, Switzerland) to monitor the movements of people (staff members, patients, and other visitors) throughout the clinic (Supplementary Figure~\zref{fig:floor-plan} in \supp). The resulting date- and timestamped movement data consisted of a person’s height, their position recorded as x-y coordinates, and a unique ID for each person's track while in the clinic. Supplementary Figure \zref{fig:tracking-examples} in \supp~shows a small, random sample of the tracks from our study. Note that individuals could contribute multiple tracks if they moved out of the range of a sensor and then returned, or if they were briefly lost because the sensor did not recognize them as a person (\eg if bending or hiding behind another person). To link these tracks, we created a R Shiny tool and manually linked tracks back together that most likely belonged to the same person (Supplementary Text~\zref{sec:setting-and-data} in \supp). We defined close contacts as other persons within a radius of $\leq$1\,m.

@@

% factors affecting quanta generation
The generation, dispersion, and survival of infectious particles in the air is subject to considerable uncertainty. Therefore, Wells\cite{Wells1955} introduced the concept of a ``quantum'' (infectious dose) to capture the stochastic nature of airborne transmission. A higher concentration of infectious quanta corresponds to a higher probability for an individual to get infected. The probability of infection is modeled with a Poisson relation and the Wells-Riley model considers multiple determinants for the generation and survival of infectious quanta. Nevertheless, transmission events are difficult to predict as the occurrence depends on multiple other factors that can broadly be categorized into (1)~environmental factors, (2)~properties of the specific pathogenic, and the (3)~infectiousness of infected and susceptibility of uninfected TB patients.

(1)~Environmental factors are somewhat considered by our model using CO$_2$ as a tracer gas to estimate outdoor air supply. The combination of clinical and video sensor data further allows the spatial modeling of quanta generation. This makes our approach unique compared to existing modeling based on the Wells-Riley equation assuming a well-mixed airspace\cite{Riley1978AJE,Rudnick2003IndoorAir}. However, the radial diffusion of quanta assumed by our model may not reflect the actual airflow at the primary care clinic. The spatial dispersion of infectious particles is often simulated with computational fluid dynamics (CFD) models\cite{Vuorinen2020SafSci,Jung2021InfectChemo,Li2021BuildEnv}. CFD models can simulate a variety of diffusion patterns, which help researchers understanding the influence of airflow. For example, considering airflow can identify poorly ventilated indoor spaces where infectious particles circulate inside the room rather than getting replaced with fresh air\cite{Li2021BuildEnv}. However, airflow is rarely considered and it may be difficult to assess in a naturally ventilated space, such as our primary care clinic, where it depends on the outdoor wind direction and which doors and windows are open.

(2)~Survival and dispersion of aerosols are also determined by the physicochemical properties of virus-laden aerosols such as particle size, viral load, and other chemical components\cite{Wang2021Science}. Currently, these properties are only indirectly incorporated in our model by modeling the quanta generation rate as a stochastic parameter. It is also challenging to consider these properties because environmental factors, in particular temperature and humidity, can affect particle generation and survival, but the effects are often unclear and depending on the pathogen\cite{Songer1967,Chan2011AdvVir,Fernstrom2013JoP,Cox1995Book,Fernstrom2013JoP,Tang2009Interface}. Survival of bacteria in aerosols decreases with temperatures $>$24$^{\circ}$C, but the effects of humidity are unclear\cite{Tang2009Interface}. On the one hand, a comparison between two studies suggested that a higher survival rate of airborne \emph{Mtb} may be attributed to lower relative humidity\cite{Loudon1969AMRRD,Lever2000LettersAppliedMicrobio}. On the other hand, airborne \emph{Mtb} was more likely to be detected in healthcare facilitates with larger relative humidity\cite{Sornboot2019IJTLD,Matuka2021IJERP}.

(3)~Previous studies show considerable variation in the generation of infectious quanta by TB infected patients\cite{Escombe2008PLoSMed,Andrews2014JID}. We model such variation by using a wide prior distribution for the quanta generation rate. Our model could be easily extended to incorporate prior information on the infectiousness of each individual TB patient. In general, only patients with active pulmonary TB are able to produce infectious droplets\cite{Rieder1999}. Infectiousness could be determined based on the most recent sputum test result. Sputum smear positive patients are more likely to transmit \emph{Mtb} than sputum smear negative patients\cite{Shaw1954ART,Brindle1993AMRRD,Grzybowski1975BIUT}. HIV status could further inform both infectiousness and susceptibility. On one hand, HIV-positive patients are more likely to have a negative sputum smear microscopy result than HIV-negative patients, due to the reduced lung cavitation and the fewer \emph{Mtb} bacilli in their sputum\cite{Brindle1993AMRRD,Telzak1997CID}. On the other hand, HIV-positive patients are less prone to control the \emph{Mtb} infection than HIV-negative patients, due to their reduced immune status\cite{Forte1992AIDS,Kwan2011CMR,Shen1988CEI}, and therefore TB disease progresses faster and is often more severe in HIV-positive TB patients compared to HIV-negative patients.

% data and other limitations
While we combined multiple data sources to allow spatiotemporal modeling of the generation, spread, and removal of infectious quanta, our data has also limitations. First, clinical attendees could not always be tracked perfectly throughout the clinic using our video sensors. We invested considerable effort to manually link individual tracking IDs probably belonging to the same attendee back together (Supplementary Text~\zref{sec:setting-and-data} in \supp), but some measurement error remains. Second, we linked video tracking and clinical patient data using the times clinical attendees were at the registration and the entry times in the clinical database. False linkages could result when a tracking ID only passed by the registration (without registering) or when multiple IDs were at the registration closely before a clinical entry. Third, to facilitate modeling, we considered the waiting room and corridor as separate, although the door between both rooms was usually open (\zref{prep:building} in \supp). Fifth, our video sensors could not cover the whole area of the entrance and we did not collect data in the care rooms and other places of the clinic. As a result, the time spent in the clinic and the exposure to infectious quanta may be underestimated for some clinical attendees, although we expect the largest exposure to be in the waiting room, which was fully covered.

>>>>

% the Wells-Riley model
The traditional Wells-Riley transmission model\cite{Riley1978AJE} is frequently used to estimate the risk of airborne transmission in a variety of indoor settings\cite{Andrews2014JID,Taylor2016IJTLD,Hella2017JInfect,Zemouri2020JDR}, including primary care clinics\cite{Zurcher2022JID,McCreesh2021BMJGlobalHealth}. The Wells-Riley model assumes a well mixed airspace, where the risk of infection is modeled as a function of the number of infectious individuals in space, the generation rate of infectious doses (called quanta), the breathing rate per person, and the outdoor air supply rate. The quanta generation rate (commonly denoted with the parameter $q$) is unknown, but can be estimated by computing the risk of infection as the ratio of diseased/susceptible cases, and then solving the Wells-Riley equation for the parameter $q$\cite{Nardell1991ARRD,Escombe2008PLoSMed}. The Wells-Riley model considers the importance of ventilation by which infectious quanta are removed from the indoor space through outdoor air exchange. While the ventilation rate is difficult to estimate, an alternative formulation of the Wells-Riley equation exists that uses indoor CO$_2$ levels as a marker of exhaled-breath exposure\cite{Rudnick2003IndoorAir}, which is related to the ventilation rate.

@@

We used an anonymous personal movement tracking system (Xovis, Zollikofen, Switzerland) to monitor the movements of people (staff members, patients, and other visitors) throughout the clinic (Supplementary Figure~\zref{fig:floor-plan} in \supp). The resulting date- and timestamped movement data consisted of a person’s height, their position recorded as x-y coordinates, and a unique ID for each person's track while in the clinic. Supplementary Figure~\zref{fig:tracking-examples} in \supp~shows a few sample tracks from our study. Note that individuals could contribute multiple tracks if they moved out of the range of a sensor and then returned, or if they were briefly lost because the sensor did not recognize them as a person (\eg if bending or hiding behind another person). To link these tracks, we created a R Shiny tool and manually linked tracks back together that most likely belonged to the same person (Supplementary Text~\zref{sec:setting-and-data} in \supp). We defined close contacts as other persons within a radius of $\leq$1\,m.

@@

% close contacts
Our spatiotemporal model can be viewed as a modified version of the original Wells-Riley transmission model\cite{Riley1978AJE}. It is motivated by previous findings that prolonged close contact is required for transmission\cite{Leung2020NatMed,Brankston2007LancetID,Narasimhan2013PulmonaryMed}. As a result, our model assigns a higher risk of infection to clinical attendees that spend more time in close proximity to infectious individuals. It could thus be used to assess the effectiveness of interventions targeting a reduction in the number of close contacts. Although such assessments only show the potential (modeled) impact of interventions, they can still provide a first estimate about how contact reductions translate into lower transmission risks in crowded indoor settings.

% more complex models and our relation
Despite our modeling efforts, transmission events remain difficult to predict. Their occurrence depends on many other factors such as airflow, the properties of the pathogen, and variation in infectiousness of the source and susceptibility of the host\cite{Wang2021Science}. We discuss these factors in more depth in Supplementary Text~\zref{sec:depth-discussion} in \supp. Importantly, there is large stream of literature considering airflow using computational fluid dynamics (CFD) models to simulate the spatial spread of infectious aerosols in indoor environments, \eg Ref. \cite{Vuorinen2020SafSci,Jung2021InfectChemo,Li2021BuildEnv}. CFD models can also be combined with the Wells-Riley model to estimate the risk of infection\cite{Yan2023BE,Qian2009BE,Li2022SOTTE}, thus are seemingly appealing to our setting as well. However, because they are computationally expensive, CFD models are often only used for experiments, short simulations, and static environments. They also require detailed knowledge about the indoor space, in particular airflow, to model particle transport. By contrast, our setting presents a dynamic environment with frequent patient movements over multiple days in a naturally ventilated space (varying airflow). Moreover, we run our model multiple times in a large number of Monte Carlo simulations to consider uncertainty in all modeling parameters and unmasked infectious TB patients.
